A Wait-free Queue with Polylogarithmic Step Complexity

### 2022-12-23 podc_paper/algorithm.tex
working on algorithm description

Each node has an infinite array called \var{blocks}.
To simplify the code, each node has an empty block in \var{blocks[0]}.
The node also has a \var{head} index which is the position in the \var{blocks} array to be used
for the next attempt to append a block to this node.

If a block contained an explicit representation of its sequences of enqueues and dequeues,
it would take $\Omega(p)$ time to construct a block, which would be much too slow for task \ref{construct}.
Instead, the block stores an implicit representation of the sequences.
We now explain how the fields of the blocks were chosen for this implicit representation. 
Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{tournament} would actually be represented, and Figure \ref{object-fields} for the definitions of the fields of blocks and tree nodes.

@@

Each block in an internal node $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  This allows us to navigate to the direct subblocks of any block easily.
Blocks also store prefix sums of the numbers of enqueues and dequeues:
the block in $v.\var{blocks}[i]$ has two fields \var{sum\sub{enq}} and \var{sum\sub{deq}}
that store the total number of enqueues and dequeues in $v.\var{blocks[1..i]}$.
These fields allow us to pinpoint the exact location of an operation among the subblocks of a given block.
For example, consider finding the $r$th enqueue in the linearization ordering.
We use the \var{sum\sub{enq}} fields of the root's blocks to do a binary search
to locate which block in the root contains the operation.
If we know which block $B$ in some node $v$ contains the enqueue,
we can use the \var{sum\sub{enq}} field again determine which child of $v$ contains the enqueue
and then to do a binary search
among the direct subblocks of $B$ in that child.
Thus, we work our way down the tree until we find a leaf block, which explicitly stores 
the enqueue.
%Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
%First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
%so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
%This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
%Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
%to determine whether the enqueue came from the left or right child of the root.
%Suppose the enqueue came from the right child $v_r$.
%Then, we know that the index of the block in $v_r$ that contains the enqueue
%is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
%We can again do a binary search within this range.
%For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
%We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other node along the path to a leaf can be done in $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps to complete task \ref{findenqueue}.
All of the information needed for this search process can be derived from the 
\eleft, \eright\ and \var{sum\sub{enq}} fields.

@@

Now that we have defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
we can easily see how to construct a new block $B$ during a \op{refresh} in $O(1)$ time.
A \op{refresh} on node $v$ reads the value $h$ of the \var{head} field of each of $v$'s children and stores 
$h-1$ in the $B.\eleft$ and $B.\eright$ fields of the new block.
Then, we can compute $B.\var{sum\sub{enq}}$ as $v.\var{left}.\var{blocks}[B.\eleft].\var{sum\sub{enq}} + v.\var{right}.\var{blocks}[B.\eright].\var{sum\sub{enq}}$.
For a block $B$ in the root, $B.\var{size}$ can be computed using the \var{size} field of the previous block,
the number of enqueues in $B$ and the number of non-null dequeues in $B$ (determined as described above).

@@

\begin{figure}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}

\Statex $\diamondsuit$ \tt{\sl{Shared Variable}}
\begin{itemize}
\item \tt{\sl{Node} root} \sf{\com\ the root of a binary tree of \tt{Node}s with one leaf for each process}
\end{itemize}

\Statex $\diamondsuit$ \tt{\sl{Local Variable}}
\begin{itemize}
\item \tt{\sl{Node} leaf} \sf{\com\ process's leaf in the tree}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{Node} left, right, parent} \textsf{\com\ children and parent pointers initialized  when creating the tree}
\item \tt{\sl{Block[0..$\infty$]} blocks} \textsf{\com\ initially \tt{blocks[0]} contains an empty block with all integer fields equal to 0}
\item \tt{\sl{int} \head} \textsf{\com\ position to attempt appending next \tt{block} to \tt{blocks}, initially 1}
\end{itemize}

@@

\begin{itemize}
  \item \tt{\sl{int} super}
  \textsf{\com\ approximate index of the block's superblock in \var{parent.blocks}}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
\begin{itemize}
    \item \tt{int} \eleft, \eright
  \textsf{\com\ indices of the last subblock of the block in the left and right child}
  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{\com\ number of enqueues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{\com\ number of dequeues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{\com\ number of enqueues in \tt{right.blocks[1..end\sub{right}]}}
  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{\com\ number of dequeues in \tt{right.blocks[1..end\sub{right}]}}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in leaves}
\begin{itemize}
  \item \tt{\sl{Object} element}
  \textsf{\com\ if the operation stored in the block is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}

    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
  \textsf{\com\ number of enqueue, dequeue operations in  leaf's \var{blocks} array up to this block (inclusive)}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following field is used only for blocks in the root}
\begin{itemize}
  \item \tt{\sl{int} \size}%
  \textsf{\com\ size of the queue after performing all operations up to the end of this block in linearization order}
\end{itemize}

@@

---------

In our case of implementing a queue, a process only needs to compute
the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific
position.  We know we can linearize operations in a block in any
order; here, we choose to put \nf{Enqueue} operations in a block
before \nf{Dequeue} operations. Consider the following operations,
where operations in a cell are concurrent. 
\begin{table}[hbtp]
\centering
\begin{small}
\begin{tabular}{c|c|c|c}
    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
  \end{tabular}
\end{small}
\caption{An execution on a queue separated by the operations in the root blocks.}
\end{table}

\noindent The \texttt{Dequeue} operations return \texttt{null, 5, 2,
  1, 3, 4, null}, respectively. 
Now, we claim that by knowing the size of the queue, we can compute
the rank of the required \texttt{Enqueue} for any non-null
\nf{Dequeue}. We apply this approach to blocks; if we store the size
of the queue after each block of operations, we can compute the index
of each \nf{Dequeue}'s result in \textsc{O}$(1)$ steps.

\begin{table}[hbtp]
\centering
\begin{small}
  \begin{tabular}{c|c|c|c|c}
    \hline &\texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
    \#\tt{Enq}s & 0 & 3 & 1 & 1 \\ \hline
    \#\tt{Deq}s & 1 & 1 & 1 & 4 \\ \hline
    Size at end & 0 & 2 & 2 & 0 \\ \hline
  \end{tabular}
\end{small}
  \caption{Augmented history of operation blocks on the queue.}
\end{table}

\noindent The size of the queue after the $b$th block in the root
could be computed as 
$$\textrm{max}\Big(\textrm{size after }(b-1)\textrm{th block} +
\#\textrm{\tt{Enqueue}s in }b\textrm{th block} -
\#\textrm{\tt{Dequeue}s in }b\textrm{th block}, 0\Big).$$ 
\noindent Moreover, the total number of non-null dequeues in blocks
$1,2,...,b$ in the root is 
$$ \sum_{i=1}^{b} \#\textrm{\tt{Enqueue}s in }i\textrm{th block} -
\textrm{size after }b\textrm{th block}.$$

Given a \texttt{Dequeue} is in block $B$, its response is the argument
of the \tt{Enqueue} whose rank is 
\begin{center}
    \#\textrm{non-null } \nf{Dequeue}s in  blocks $1,2,...,b-1 +  $index of the \nf{Dequeue} among B\textrm{'s} \nf{Dequeue}
\end{center}
if \vspace{-3em}
\begin{center}
    $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} $\\$- \textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$.
\end{center}
Otherwise, the response would be \nf{null}.

>>>>

Each node has an infinite array called \var{blocks}.
To simplify the code, each node has an empty block in \var{blocks[0]}.
Each node also has a \var{head} index that stores the position in the \var{blocks} array to be used
for the next attempt to append a block to this node.

If a block contained an explicit representation of its sequences of enqueues and dequeues,
it would take $\Omega(p)$ time to construct a block, which would be too slow for task \ref{construct}.
Instead, the block stores an implicit representation of the sequences.
We now explain how we designed the fields for this implicit representation. 
Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{tournament} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of blocks and tree nodes.

@@

\here{If space permits, we might want to add some examples in the following paragraphs that refer back to Figure \ref{implicit}.}

Each block in an internal node $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  This allows us to navigate to the direct subblocks of any block easily.
Blocks also store prefix sums of the numbers of enqueues and dequeues:
the block in $v.\var{blocks}[i]$ has two fields \var{sum\sub{enq}} and \var{sum\sub{deq}}
that store the total number of enqueues and dequeues in $v.\var{blocks[1..i]}$.
These fields allow us to pinpoint the exact location of an operation among the subblocks of a given block.
For example, consider finding the $r$th enqueue in the linearization ordering.
We use the \var{sum\sub{enq}} fields of the root's blocks to do a binary search
to locate which block in the root contains the operation.
If we know which block $B$ in some node $v$ contains the enqueue,
we can use the \var{sum\sub{enq}} field again to determine which child of $v$ contains the enqueue
and then to do a binary search
among the direct subblocks of $B$ in that child.
Thus, we work our way down the tree until we find a leaf block, which explicitly stores 
the enqueue.
%Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
%First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
%so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
%This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
%Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
%to determine whether the enqueue came from the left or right child of the root.
%Suppose the enqueue came from the right child $v_r$.
%Then, we know that the index of the block in $v_r$ that contains the enqueue
%is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
%We can again do a binary search within this range.
%For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
%We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other node along the path to a leaf can be done in $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps to complete task \ref{findenqueue}.
All of the information needed for this search process can be derived from the 
\eleft, \eright\ and \var{sum\sub{enq}} fields.

@@

Now that we have defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
we can easily see how to construct a new block $B$ during a \op{refresh} in $O(1)$ time.
A \op{refresh} on node $v$ reads the value $h$ of the \var{head} field of each of $v$'s children and stores 
$h-1$ in the $B.\eleft$ and $B.\eright$ fields of the new block.
Then, we can compute $B.\var{sum\sub{enq}}$ as $v.\var{left}.\var{blocks}[B.\eleft].\var{sum\sub{enq}} + v.\var{right}.\var{blocks}[B.\eright].\var{sum\sub{enq}}$.
For a block $B$ in the root, $B.\var{size}$ can be computed using the \var{size} field of the previous block $B'$ and
the number of enqueues and dequeues in $B$:
$B.\var{size} = \max(0, B'.\var{size} + (B.\var{sum\sub{enq}}-B'.\var{sum\sub{enq}}) - (B.\var{sum\sub{deq}} - B'.\var{sum\sub{deq}}))$.

@@

\begin{figure}
\here{Does \var{leaf} have to be shared for helping when doing GC?}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}
\Statex $\diamondsuit$ \tt{\sl{Shared variable}}
\begin{itemize}
\item \tt{\sl{Node} root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process}
\end{itemize}

\Statex $\diamondsuit$ \tt{\sl{Thread-local variable}} 
\begin{itemize}
\item \tt{\sl{Node} leaf} \Comment{process's leaf in the tree}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{Node} left, right, parent} \Comment{children and parent pointers initialized  when creating the tree}
\item \tt{\sl{Block[0..$\infty$]} blocks} \Comment{initially \tt{blocks[0]} contains an empty block with all integer fields equal to 0}
\item \tt{\sl{int} \head} \Comment{position to attempt appending next \tt{block} to \tt{blocks}, initially 1}
\end{itemize}

@@

\begin{itemize}
  	\item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
  		\Comment{number of enqueue, dequeue operations in \var{blocks} array up to this block (inclusive)}
  	\item \tt{\sl{int} super}
  		\Comment{approximate index of the block's superblock in \var{parent.blocks}}
	\item[\com] Blocks in internal nodes have the following additional fields
	\item \tt{int} \eleft, \eright
  		\Comment{index of last direct subblock in the left and right child}
  	\item[\com] Blocks in leaf nodes have the following additional field
  	\item \tt{\sl{Object} element}
  		\Comment{if the block's operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
	\item[\com] Blocks in the root node have the following additional field
	\item \tt{\sl{int} \size}%
  		\Comment{size of the queue after performing all operations up to the end of this block}
\end{itemize}

%\Statex {\com\  Blocks in internal nodes have the following additional fields}
%\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
%\begin{itemize}
%	\item \tt{int} \eleft, \eright
%  		\Comment{index of last direct subblock in the left and right child}
%\end{itemize}

%\Statex {\com\ Blocks in leaf nodes have the following additional field}
%\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional field is used only for blocks in leaves}
%\begin{itemize}
%  \item \tt{\sl{Object} element}
%  \Comment{if the block's operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
%\end{itemize}

%\Statex {\com\ Blocks in the root node have the following additional field}
%\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following additional field is used only for blocks in the root}
%\begin{itemize}
%  \item \tt{\sl{int} \size}%
%  \Comment{size of the queue after performing all operations up to the end of this block}
%\end{itemize}

@@

### 2022-12-23 podc_paper/introduction.tex
Update introduction.tex
working on algorithm description

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
IT uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

>>>>

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21} \here{try to find an older, more canonical reference}),
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
It uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.
